Thesis: draft

### 2021-05-31 draft.tex
final version?

The past decade social media has become ever more important in our daily lives. Nowadays most of us have, beside a real-world life, a virtual life on social media. This increased engagement on social media may have an impact on the formation of opinions, not only on the individual scale, but also on a broader, societal scale. \\
\newline
It is known that individuals not only form their opinions through self-reflection, but also through interactions with other people and with their surroundings \cite{Perra2019}. The broad range of interactions people undergo on social media with people from all over the world can thus not be underestimated in the process of opinion formation. The understanding of the role of social media on the emergence of polarization and extremism in society is of uttermost importance.\\
\newline
Some people change their opinions more easily than others, that is, once in a while you encounter somebody that is stubborn and persistent towards its own opinion. It is important to understand the effect of individual resistance on the formation and evolution of opinions. When designing models of opinion dynamics on social networks, this individual stubbornness should not be ignored.\\
\newline
One important feature of opinion dynamics on social media is the use of algorithmic personalization. Social media allow for simultaneous or asynchronous interactions between people without geographical constraints and thus allow for the spread of information at a faster pace and at an unprecedented scale \cite{Perra2019}. People are, however, still bound by temporal and cognitive constraints. In order to assure a more pleasant and convenient experience on their platforms, social media companies use algorithms to order and filter the posts on an individual's time-line according to what might be relevant to that individual \cite{Perra2019}. It is of importance to know the effect of these filtering algorithms on the evolution of opinions in the population.\\  
\newline
Opinion dynamics models have two important layers. On one hand, we introduce social networks to describe the underlying structure of social interactions; on the other hand, we need appropriate models to reproduce the opinion dynamics and opinion formation processes found in real life.
\newpage
\noindent
Section \ref{statPhys} introduces concepts such as statistical and social physics. Section \ref{complNet} gives an introduction to the first layer, that of complex networks. The second layer, the opinion dynamics, is introduced in Section \ref{Opinion}. Finally, in Section \ref{goal}, the hypotheses and goals of this thesis are summarized and discussed.

@@

An important tool in the study of opinion dynamics in large groups of people is the use of statistical physics. This branch of physics offers a framework that relates microscopic properties of atoms and molecules to macroscopic observed behavior and has a wide range of applications inside and outside physics. The observation that large number of people display collective, `macroscopic' behavior begged for the use of the concepts and insights developed in statistical physics \cite{Sirbu2016}. The application of the theory of statistical physics on social phenomena is referred to as social physics \cite{Castellano2009}\cite{Galam2008}\cite{Galam1982}\cite{Stauffer2012}. The `microscopic' constituents are now individual humans who interact with a limited number of other individuals and in that way form complex, `macroscopic' groups such as human societies. These `macroscopic' groups display regularities, transitions from disorder to order, the emergence of consensus, universality \cite{Buchanan2007}. The statistical physics approach of these social systems tries to explain the regularities at large scale as collective effects of the interactions among individuals \cite{Sirbu2016}.\\
\newline
The statistical approach of social phenomena has one difficulty compared to the description of gases and other physical phenomena. In physics, the microscopic constituents are atoms or molecules, whose dynamics and interactions with each other are often well known and can be put in relatively simple laws with only a few parameters \cite{Castellano2009}. Humans, on the other hand, are far from simple entities and the dynamics of a single individual are far from known \cite{Castellano2009}. Even if the dynamics of individuals would be well know, the interactions between them would still be complicated and it would be hard to describe them with simple laws and only a few parameters \cite{Castellano2009}. Hence, extracting useful and qualitative information out of social models is a challenging task \cite{Castellano2009}. The use of statistical physics may, however, offer a way out. In statistical physics it is found that different systems behave qualitatively the same and are governed by the same laws near the critical point \cite{Hu2018}\cite{Kadanoff2010}. This lead to the concept of universality: many different microscopic problems have the same macroscopic manifestation \cite{Gug2015}\cite{Kadanoff2015}. This idea may be extended to social systems \cite{Hu2018}: we should not take the unique character of every individual into account, instead these individual characteristics can be incorporated in a sort of averaged way \cite{Gug2015}. One can then proceed to build models of social behavior where only the most important properties of single individuals and their interactions are taken into account \cite{Castellano2009}. An important factor herein is the comparison with empirical data to investigate whether the trends seen in real life are compatible with the outputs generated by the models \cite{Castellano2009}\cite{Pawel}. This interplay between theory, model, experiment and real-life data is not only an important factor in social physics, but also in physical research and science in general.

@@

The formation of an individual's opinion is the result of the interplay of many factors such as self-reflection, peer pressure, the individual's personality (e.g. stubbornness) and the information someone is exposed to. Opinion dynamics models should try to capture these complex processes in a simplified way. Several models have been developed ranging from simple binary models to more complex, continuous approaches \cite{Sirbu2016}.\\
\newline
The basic idea of all opinion dynamics models is that the nodes or actors in a social network have a variable that represents their opinion and that is updated according to some predefined rules. These models are a simplification of real-world opinion dynamics. It is however shown that they display aspects of real opinion formation such as agreement, transitions between order (consensus) and disorder (fragmentation), polarization and formation of echo chambers (clusters of people with the same opinion) \cite{Sirbu2016}. \\
\newline
The opinion of the actors in the model can be either discrete or continuous. This thesis deals with models where each actor can have one of two opinions, A or B, and thus only deals with the case of discrete opinions. This might come across as a simplification of the real-world complexity of opinion formation, but also in real life people often have to choose between two competing opinions (e.g. republicans or democrats, renting or buying a house). \\
\newline
The rules that determine how an actor updates his or her opinion can include different aspects. They can, for example, be as simple as a majority rule (i.e. the majority model: if the majority of your neighbors have a certain opinion, you adopt that opinion as well) or can include a probabilistic way of updating. An overview of possible opinion dynamics models is given in Section \ref{Bin} and Section \ref{Con}. One can also include a resistance parameter that determines the hesitation of an actor to change to a new opinion or a parameter that determines the influence of an actor on others. Furthermore, the actors can have a parameter that determines whether they are active or not. The introduction of stubborn actors will be discussed in Section \ref{stubb}; the activation mechanism of actors is discussed in Section \ref{actMech}.\\
\newline
Since this thesis deals with the particular case of opinion dynamics on on-line social platforms, it is important to define filtering algorithms. The filtering algorithms that social media companies use, are corporate secrets, but they typically have three main principles of content curation: popularity, semantic and collaborative filtering \cite{Perra2019}. Popularity filtering refers to the practice of promoting content that is popular across the platform; semantic filtering means that post similar to previous consumed posts are recommended and collaborative filtering suggests posts that are similar to the ones our friends consume \cite{Perra2019}. In Section \ref{filter} and Section \ref{modelThesis} a detailed description of the filtering algorithms used in this thesis will be given.
\newpage
\section{This thesis: hypotheses and goals}\label{goal}

@@

Another concept that is closely related to the degree of a node is assortative and disassortative networks. The notion of assortativity and disassortativity has to do with degree correlations. \\
\newline
The joint degree distribution $P(k, k')$ is the probability that an arbitrary edge connects a node with degree $k$ to a node with degree $k'$ \cite{F.Costa2007}. The conditional probability $P(k|k')$, which is another way to express dependencies between node degrees, is the probability that an arbitrary neighbor of a node with degree $k$ has a degree $k'$ 
\begin{equation}
	P(k|k') = \frac{\left<k\right>P(k, k')}{kP(k)} \ ,
\end{equation}
where a neighbor of a node $i$ is an adjacent node of node $i$, the meaning of adjacent nodes is given in Section \ref{connect} \cite{F.Costa2007}. The average degree of the nearest neighbors of nodes with degree $k$ can be computed as \cite{F.Costa2007}
\begin{equation}
	k_{nn}(k) = \sum_{k'} k'P(k'|k) \ .
\end{equation}
If there are no degree correlations, $k_{nn}$ is independent of $k$. Such a network is called a neutral network \cite{F.Costa2007}. If $k_{nn}(k)$ is an increasing function of $k$, nodes with a high degree tend to connect to nodes with a high degree and the network is classified as assortative. If instead $k_{nn}(k)$ is a decreasing function of $k$, nodes with a high degree tend to connect to nodes with a low degree and the network is called disassortative \cite{F.Costa2007}.\\
\newline
Assortativity and disassortativity are most often applied to the degree of the nodes in the network, which is referred to as degree assortativity \cite{Noldus2015}. Assortativity may, however, also be applied to other characteristics of the nodes \cite{Noldus2015}\cite{Thed2014}. In this way, an assortative network is a network where nodes tend to connect to other nodes with similar properties as themselves, whereas in a disassortative network nodes tend to connect to nodes with different properties \cite{Thed2014}.

@@

Several measurements quantify the notion of length and distance in a network. A path is a sequence of distinct nodes, such that adjacent nodes in the sequence are adjacent nodes in the network \cite{Goddard2010}. The length of a path is defined as the number of edges in the path \cite{Goddard2010}. A geodesic path, often called the shortest path, between two nodes is the path between two nodes $i$ and $j$ with minimum length (the geodesic path is not necessarily unique) \cite{F.Costa2007}. The length of the geodesic path between the two nodes $i$ and $j$ is called the geodesic distance or shortest distance $d_{ij}$ \cite{F.Costa2007}. Geodesic distance is often just called distance \cite{Goddard2010}. The notion of distance enables to define several global network measurements such as average distance, diameter and radius.\\  
\newline
\textbf{Average distance}\\
\newline
The average distance is the mean value of the distance $d_{ij}$ \cite{F.Costa2007}
\begin{equation}\label{avdist}
	\left<d\right> = \frac{1}{N(N-1)} \sum_{i \neq j} d_{ij} \ .
\end{equation}
This definition, however, diverges if there are unconnected nodes in the network (unconnected nodes have, by definition, a distance equal to infinity) \cite{F.Costa2007}. This problem could be avoided by only including connected pairs in the sum, but this introduces a distortion for networks with a lot of unconnected pairs of nodes. These networks will then have a small average distance, which is only expected for networks with lots of connections \cite{F.Costa2007}. Instead, another definition of average distance can be used: the global efficiency $E$ \cite{F.Costa2007}
\begin{equation}
	E = \frac{1}{N(N-1)} \sum_{i \neq j} \frac{1}{d_{ij}} \ .
\end{equation}
This measurement quantifies the efficiency of the network in sending information between nodes \cite{F.Costa2007}. This assumes that the efficiency of sending information between two nodes is reciprocal to their distance. The harmonic mean of the geodesic distances is now defined as the reciprocal of the global efficiency \cite{F.Costa2007}
\begin{equation}
 	h = \frac{1}{E} \ .
\end{equation}
\newline
\textbf{Diameter and radius}\\
\newline
The diameter $diam(G)$ of a network $G$ is the maximum distance between any pair of nodes \cite{Goddard2010}. The eccentricity of a node is the maximum distance from that node to any other node. The radius $rad(G)$ is the minimum eccentricity among all nodes of $G$. Note that the diameter is the maximum eccentricity among all nodes in $G$ \cite{Goddard2010}.
\newpage
\noindent
\subsection{Clustering coefficient and transitivity}

@@

The modularity of a network with $N$ nodes and $L$ edges and partitioned into $n_c$ communities is calculated as follows. Each community has $N_c$ nodes and $L_c$ edges ($c = 1,..., n_c$) \cite{Albert2016}. The modularity of a community, $Q_c$, is calculated by measuring the difference between the network's real wiring diagram (which is given by the adjacency matrix $A_{ij}$) and the expected number of edges between $i$ and $j$ if the network is randomly wired (where $i$ and $j$ are two nodes that lie in the same community $C_c$) \cite{Albert2016}. This last quantity is given by $p_{ij}$, which is obtained by randomizing the original network while keeping the expected degree of each node unchanged \cite{Albert2016}. Hence, the modularity of a community $Q_c$ becomes

@@

The Erd\H{o}s-R\'{e}nyi model is used as a benchmark model. We focus on two network topologies that may impact the formation of echo chambers: high clustering and community structure. It is therefore necessary to have a model that lacks both of these features to compare the results with. The Erd\H{o}s-R\'{e}nyi model is one of the most simple random network models that does not display both of these features and is hence the appropriate comparison model. It is also used in the regular stochastic block model to construct the constituent communities.\\

@@

The stochastic block model (SBM) is a random network model that is able to produce networks containing community structure. The stochastic block model originates in social sciences and was developed to describe group structures in friendship networks \cite{Funke2019}. The idea behind the development of this model is that often in real networks, there are groups of nodes that are more densely connected with each other than with the rest of the network \cite{F.Costa2007}. Hence, these networks posses a modular structure and this is a structure that is easily modeled by the stochastic block model. The stochastic block model is also often used as a benchmark for community detection algorithms \cite{Abbe2018}.\\

@@

The stochastic block model can be fitted to data \cite{Clauset2017}. Given a choice of the number of communities $k$ and an observed network $G$, the stochastic block model can be used to infer the latent community structure and the edge probability matrix $M$ \cite{Clauset2017}. One of the shortcomings of fitting the SBM to infer the community structure, is that real-world networks often have a heterogeneous or power-law degree distribution. Since the SBM is made of random networks inside each group and random bipartite networks between each pair of groups, the degree distribution of the full network is always a mixture of Poisson distributions \cite{Clauset2017}. The consequence of this is that if a network possesses a skewed degree distribution, then fitting the SBM to the data will most likely give poor results \cite{Karrer2011}. The reason behind this is one of the main assumptions of the stochastic block model: the edge probabilities only depend on which group the node belongs to \cite{Clauset2017}. In other words, nodes that belong to the same group are stochastically equivalent, meaning that they have equivalent connectivity patterns to other nodes \cite{Clauset2017}. A network with a skewed degree distribution violates this assumption of edge independence \cite{Clauset2017}. To say it in the words of Karrer and Newman: ``Just as the fitting of a straight line to intrinsically curved data is likely to miss important features of the data, so a fit of the simple stochastic block model to the structure of a complex network is likely to miss much [...].'' (Karrer and Newman, p.1, \cite{Karrer2011}). Karrer and Newman proposed a variation of the SBM, the degree corrected SBM, that deals with this limitation. We will not discuss their variation, since that would lead us too far astray; interested readers can find more information in \cite{Karrer2011}. \\

@@

We will use two real-world networks. The first real-world network is an on-line friendship network among Finnish users of the social music listening and sharing website Last.fm (data from $2006$). Only the largest connected component of this on-line social network is included \cite{ICON}. This data set was originally used in \cite{Toivonen2009}. This on-line social network has size $N = 8003$; number of edges: $16824$; $Q = 0.79$; $\left<cc\right> = 0.31$; $\left<k\right> = 4.2$. The degree distribution is heterogeneous and is best fitted by a log-normal

@@

The second real-world network is the giant component of the on-line social network of users of the Pretty-Good-Privacy (PGP) algorithm for secure information interchange (data from $2004$) \cite{Boguna2004}\cite{ICON}. This network has $N = 10680$; number of edges: $24340$; $Q = 0.85$; $\left<cc\right> = 0.27$; $\left<k\right> = 4.56$. The network has a heterogeneous degree distribution, which is best fitted by two power-laws

@@

\begin{figure}[H]
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{}
    \includegraphics[width=8cm, height=7cm]{lastfm_degree_distribution_8x7.png}
    \label{deg_distr_lastfm}
  \end{subfigure}
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{}
    \includegraphics[width=8cm, height=7cm]{PGP_degree_distribution_8x7.png}
    \label{deg_distr_pgp}
  \end{subfigure}
  \captionsetup{format=plain}
  \caption[Degree distribution $P(k)$ of the two real-world networks used in this thesis.]{Degree distribution $P(k)$ of the two real-world networks used in this thesis. \textbf{(a)} Last.fm on-line social network with a log-normal fit ($\mu = 1.008$, $\sigma = 0.95$); \textbf{(b)} PGP on-line trust network with two power-law fits ($a = 0.39$, $\gamma = 1.54$ for $k < 20$ and $a = 5.59$, $\gamma = 2.49$ for $k \geqslant 20$). Curves are fitted to the data with \textsc{scipy.optimize.curve$\_$fit}.}
\label{deg_distr_real_network}
\end{figure}
\noindent
One of the decisive properties of the real-world networks that are used, is their size. The main, and actually only, reason for this is the computation time. If the networks are too big, it takes weeks, sometimes even months before there would be results. The size of the network models that are implemented, is around a thousand nodes. For the real-world networks, we also use networks with a size of magnitude of thousand to ten-thousand. This benefits the computation time.

\section{Temporal networks}

@@

Binary models are models where each actor can have one of two opinions A or B. A binary model reflects real-life where there exist two competing opinions, such as a vote in an election system with two competing candidates or choosing between Windows and Linux \cite{Nguyen2020}.\\
\newline
In this section, the most important binary models will be discussed.

@@

Most of the standard opinion dynamics models use a static network in which actors are always active and exchanging opinions. In these models the opinion dynamics might go as follows: each time step a random actor is selected after which its opinion is updated according to the opinion dynamics model that is implemented. Another possibility is updating all actors at the same time during each time step. \\
\newline
It might be more realistic if actors become active/inactive over time. This would be a better representation of social media platforms, where users are active for some periods of time and then become inactive again. This activation mechanism of the actors gives the network a temporal character. The underlying network structure does not change (each actor has a fixed number of neighbors that does not change over time; no new friends are made or, conversely, no old friends are being unfriended). However, by making actors active and inactive the exchange of information between actors becomes time dependent on which actors are active or not. Hence, this does affect the network structure in some way, because if an actor, for example, does not become active for a long period of time, it does not participate in the exchange of information and the network structure behaves like if this actor is absent. This may affect the dynamics, since it may seem as if the network has a different average degree and some properties like connectivity and clustering may behave differently. \\
\newline
In this thesis, actors are made active with some probability $p_i$ every time step (thus, every time step, each actor becomes active with a probability $p_i$). The activation probability is set equal to $p = 0.1$, which is constant for all the actors. This constant activation probability is chosen for simplicity.

>>>>

The past decade social media has become ever more important in our daily lives. Nowadays most of us have, beside a real-world life, a virtual life on social media. This increased engagement on social media may have an impact on the formation of opinions, not only on the individual scale, but also on a broader, societal scale. \\
\newline
It is known that individuals not only form their opinions through self-reflection, but also through interactions with other people and with their surroundings \cite{Perra2019}. The broad range of interactions people undergo on social media with people from all over the world can thus not be underestimated in the process of opinion formation. The understanding of the role of social media on the emergence of polarization and extremism in society is of uttermost importance.\\
\newline
Some people change their opinions more easily than others, that is, once in a while you encounter somebody that is stubborn and persistent towards its own opinion. It is important to understand the effect of individual resistance on the formation and evolution of opinions. When designing models of opinion dynamics on social networks, this individual stubbornness should not be ignored.\\
\newline
One important aspect of opinion dynamics on social media is the use of algorithmic personalization. Social media allow for simultaneous or asynchronous interactions between people without geographical constraints and thus allow for the spread of information at a faster pace and at an unprecedented scale \cite{Perra2019}. People are, however, still bound by temporal and cognitive constraints. In order to assure a more pleasant and convenient experience on their platforms, social media companies use algorithms to order and filter the posts on an individual's time-line according to what might be relevant to that individual \cite{Perra2019}. It is of importance to know the effect of these filtering algorithms on the evolution of opinions in the population.\\  
\newline
Opinion dynamics models have two important layers. On one hand, we introduce social networks to describe the underlying structure of social interactions; on the other hand, we need appropriate models to reproduce the opinion dynamics and opinion formation processes found in real life.
\newpage
\noindent
Section \ref{statPhys} introduces concepts such as statistical and social physics. Section \ref{complNet} gives an introduction to the first layer, that of complex networks. The second layer, the opinion dynamics, is introduced in Section \ref{Opinion}. Finally, in Section \ref{goal}, the hypotheses and goals of this thesis are summarized and discussed.

@@

An important tool in the study of opinion dynamics in large groups of people is the use of statistical physics. This branch of physics offers a framework that relates microscopic properties of atoms and molecules to macroscopic observed behavior and has a wide range of applications inside and outside physics. The observation that large number of people display collective, `macroscopic' behavior begged for the use of the concepts and insights developed in statistical physics \cite{Sirbu2016}. The application of the theory of statistical physics on social phenomena is referred to as social physics or sociophysics \cite{Castellano2009}\cite{Galam2008}\cite{Galam1982}\cite{Stauffer2012}. The `microscopic' constituents are now individual humans who interact with a limited number of other individuals and in that way form complex, `macroscopic' groups such as human societies. These `macroscopic' groups display regularities, transitions from disorder to order, the emergence of consensus, universality \cite{Buchanan2007}. The statistical physics approach of these social systems tries to explain the regularities at large scale as collective effects of the interactions among individuals \cite{Sirbu2016}.\\
\newline
The statistical approach of social phenomena has one difficulty compared to the description of gases and other physical phenomena. In physics, the microscopic constituents are atoms or molecules, whose dynamics and interactions with each other are often well known and can be put in relatively simple laws with only a few parameters \cite{Castellano2009}. Humans, on the other hand, are far from simple entities and the dynamics of a single individual are far from known \cite{Castellano2009}. Even if the dynamics of individuals would be well know, the interactions between them would still be complicated and it would be hard to describe them with simple laws and only a few parameters \cite{Castellano2009}. Hence, extracting useful and qualitative information out of social models is a challenging task \cite{Castellano2009}. The use of statistical physics may, however, offer a way out. In statistical physics it is found that different systems behave qualitatively the same and are governed by the same laws near the critical point \cite{Hu2018}\cite{Kadanoff2010}. This lead to the concept of universality: many different microscopic problems have the same macroscopic manifestation \cite{Gug2015}\cite{Kadanoff2015}. This idea may be extended to social systems \cite{Hu2018}: we should not take the unique character of every individual into account, instead these individual characteristics can be incorporated in a sort of averaged way \cite{Gug2015}. One can then proceed to build models of social behavior where only the most important properties of single individuals and their interactions are taken into account \cite{Castellano2009}. An important factor herein is the comparison with empirical data to investigate whether the trends seen in real life are compatible with the outputs generated by the models \cite{Castellano2009}\cite{Pawel}. This interplay between theory, model, experiment and real-life data is not only an important factor in social physics, but also in physical research and science in general.

@@

The formation of an individual's opinion is the result of the interplay of factors such as self-reflection, peer pressure, the individual's personality (e.g. stubbornness) and the information someone is exposed to. Opinion dynamics models should try to capture these complex processes in a simplified way. Several models have been developed ranging from simple binary models to more complex, continuous approaches \cite{Sirbu2016}.\\
\newline
The basic idea of all opinion dynamics models is that the nodes or actors in a social network have a variable that represents their opinion and that is updated according to some predefined rules. These models are a simplification of real-world opinion dynamics. It is however shown that they display aspects of real opinion formation such as agreement, transitions between order (consensus) and disorder (fragmentation), polarization and formation of echo chambers (clusters of people with the same opinion) \cite{Sirbu2016}. \\
\newline
The opinion of the actors in the model can be either discrete or continuous. This thesis deals with models where each actor can have one of two opinions, A or B, and thus only deals with the case of discrete opinions. This might come across as a simplification of the real-world complexity of opinion formation, but also in real life people often have to choose between two competing opinions (e.g. republicans or democrats, renting or buying a house). \\
\newline
The rules that determine how an actor updates his or her opinion can include different aspects. They can, for example, be as simple as a majority rule (i.e. the majority model: if the majority of your neighbors have a certain opinion, you adopt that opinion as well) or can include a probabilistic way of updating. An overview of possible opinion dynamics models is given in Section \ref{Bin} and Section \ref{Con}. One can also include a resistance parameter that determines the hesitation of an actor to change to a new opinion or a parameter that determines the influence of an actor on others. Furthermore, the actors can have a parameter that determines whether they are active or not. The introduction of stubborn actors will be discussed in Section \ref{stubb}; the activation mechanism of actors is discussed in Section \ref{actMech}.\\
\newline
Since this thesis deals with the particular case of opinion dynamics on on-line social platforms, it is important to define filtering algorithms. The filtering algorithms that social media companies use, are corporate secrets, but typically have three main principles of content curation: popularity, semantic and collaborative filtering \cite{Perra2019}. Popularity filtering refers to the practice of promoting content that is popular across the platform; semantic filtering means that post similar to previous consumed posts are recommended and collaborative filtering suggests posts that are similar to the ones our friends consume \cite{Perra2019}. In Section \ref{filter} and Section \ref{modelThesis} a detailed description of the filtering algorithms used in this thesis will be given.
\newpage
\section{This thesis: hypotheses and goals}\label{goal}

@@

Another concept that is closely related to the degree of a node is assortative and disassortative networks. Assortativity and disassortativity have to do with degree correlations. \\
\newline
The joint degree distribution $P(k, k')$ is the probability that an arbitrary edge connects a node with degree $k$ to a node with degree $k'$ \cite{F.Costa2007}. The conditional probability $P(k|k')$, which is another way to express dependencies between node degrees, is the probability that an arbitrary neighbor of a node with degree $k$ has a degree $k'$ 
\begin{equation}
	P(k|k') = \frac{\left<k\right>P(k, k')}{kP(k)} \ ,
\end{equation}
where a neighbor of a node $i$ is an adjacent node of node $i$, the meaning of adjacent nodes is given in Section \ref{connect} \cite{F.Costa2007}. The average degree of the nearest neighbors of nodes with degree $k$ can be computed as \cite{F.Costa2007}
\begin{equation}
	k_{nn}(k) = \sum_{k'} k'P(k'|k) \ .
\end{equation}
If there are no degree correlations, $k_{nn}$ is independent of $k$. Such a network is called a neutral network \cite{F.Costa2007}. If $k_{nn}(k)$ is an increasing function of $k$, nodes with a high degree tend to connect to nodes with a high degree and the network is classified as assortative. If instead $k_{nn}(k)$ is a decreasing function of $k$, nodes with a high degree tend to connect to nodes with a low degree and the network is called disassortative \cite{F.Costa2007}.\\
\newline
Assortativity and disassortativity are most often applied to the degree of the nodes in the network, which is referred to as degree assortativity \cite{Noldus2015}. Assortativity may, however, also be applied to other characteristics of the nodes \cite{Noldus2015}\cite{Thed2014}. In this way, an assortative network is a network where nodes tend to connect to other nodes with similar properties as themselves, whereas in a disassortative network nodes tend to connect to nodes with different properties \cite{Thed2014}.

@@

Several measurements quantify length and distance in a network. A path is a sequence of distinct nodes, such that adjacent nodes in the sequence are adjacent nodes in the network \cite{Goddard2010}. The length of a path is defined as the number of edges in the path \cite{Goddard2010}. A geodesic path, often called the shortest path, between two nodes is the path between two nodes $i$ and $j$ with minimum length (the geodesic path is not necessarily unique) \cite{F.Costa2007}. The length of the geodesic path between the two nodes $i$ and $j$ is called the geodesic distance or shortest distance $d_{ij}$ \cite{F.Costa2007}. Geodesic distance is often just called distance \cite{Goddard2010}. The notion of distance enables to define several global network measurements such as average distance, diameter and radius.\\  
\newline
\textbf{Average distance}\\
\newline
The average distance is the mean value of the geodesic $d_{ij}$ \cite{F.Costa2007}
\begin{equation}\label{avdist}
	\left<d\right> = \frac{1}{N(N-1)} \sum_{i \neq j} d_{ij} \ .
\end{equation}
This definition diverges if there are unconnected nodes in the network (unconnected nodes have, by definition, a distance equal to infinity) \cite{F.Costa2007}. This can be avoided by only including connected pairs in the sum, but this introduces a distortion for networks with a lot of unconnected pairs of nodes. These networks will then have a small average distance, which is only expected for networks with lots of connections \cite{F.Costa2007}. Instead, another definition of average distance can be used: the global efficiency $E$ \cite{F.Costa2007}
\begin{equation}
	E = \frac{1}{N(N-1)} \sum_{i \neq j} \frac{1}{d_{ij}} \ .
\end{equation}
This measurement quantifies the efficiency of the network in sending information between nodes \cite{F.Costa2007}. This assumes that the efficiency of sending information between two nodes is reciprocal to their distance. The harmonic mean of the geodesic distances is now defined as the reciprocal of the global efficiency \cite{F.Costa2007}
\begin{equation}
 	h = \frac{1}{E} \ .
\end{equation}
\newline
\textbf{Diameter and radius}\\
\newline
The diameter $diam(G)$ of a network $G$ is the maximum distance between any pair of nodes \cite{Goddard2010}. The eccentricity of a node is the maximum distance from that node to any other node. The radius $rad(G)$ is the minimum eccentricity among all nodes of $G$. Note that the diameter is the maximum eccentricity among all nodes in $G$ \cite{Goddard2010}.
\newpage
\noindent
\subsection{Clustering coefficient and transitivity}

@@

The modularity of a network with $N$ nodes and $L$ edges and partitioned into $n_c$ communities is calculated as follows. Each community has $N_c$ nodes and $L_c$ edges ($c = 1, ..., n_c$) \cite{Albert2016}. The modularity of a community, $Q_c$, is calculated by measuring the difference between the network's real wiring diagram (which is given by the adjacency matrix $A_{ij}$) and the expected number of edges between $i$ and $j$ if the network is randomly wired (where $i$ and $j$ are two nodes that lie in the same community $C_c$) \cite{Albert2016}. This last quantity is given by $p_{ij}$, which is obtained by randomizing the original network while keeping the expected degree of each node unchanged \cite{Albert2016}. Hence, the modularity of a community $Q_c$ becomes

@@

The Erd\H{o}s-R\'{e}nyi model is used as a benchmark model. We focus on two network topologies that may impact the formation of echo chambers: high clustering and community structure. It is therefore necessary to have a model that lacks both of these topologies to compare the results with. The Erd\H{o}s-R\'{e}nyi model is one of the simplest random network models that does not display both of these topologies and is hence the appropriate comparison model. It is also used in the regular stochastic block model to construct the constituent communities.\\

@@

The stochastic block model (SBM) is a random network model that is able to produce networks containing community structure. The stochastic block model originates in social sciences and was developed to describe group structures in friendship networks \cite{Funke2019}. The idea behind the development of this model is that often in real networks, there are groups of nodes that are more densely connected with each other than with the rest of the network \cite{F.Costa2007}. Hence, these networks posses a modular structure and this is a structure that can be modeled by the stochastic block model. The stochastic block model is also often used as a benchmark for community detection algorithms \cite{Abbe2018}.\\

@@

The stochastic block model can be fitted to data \cite{Clauset2017}. Given a choice of the number of communities $k$ and an observed network $G$, the stochastic block model can be used to infer the latent community structure and the edge probability matrix $M$ \cite{Clauset2017}. One of the shortcomings of fitting the SBM to infer the community structure, is that real-world networks often have a heterogeneous or power-law degree distribution. Since the SBM is made of random networks inside each group and random bipartite networks between each pair of groups, the degree distribution of the full network is always a mixture of Poisson distributions \cite{Clauset2017}. The consequence is that if a network possesses a skewed degree distribution, then fitting the SBM to the data will most likely give poor results \cite{Karrer2011}. The reason behind this is one of the main assumptions of the stochastic block model: the edge probabilities only depend on which group the node belongs to \cite{Clauset2017}. In other words, nodes that belong to the same group are stochastically equivalent, meaning that they have equivalent connectivity patterns to other nodes \cite{Clauset2017}. A network with a skewed degree distribution violates this assumption of edge independence \cite{Clauset2017}. To say it in the words of Karrer and Newman: ``Just as the fitting of a straight line to intrinsically curved data is likely to miss important features of the data, so a fit of the simple stochastic block model to the structure of a complex network is likely to miss much [...].'' (Karrer and Newman, p.1, \cite{Karrer2011}). Karrer and Newman proposed a variation of the SBM, the degree corrected SBM, that deals with this limitation. We will not discuss their variation, since that would lead us too far astray; interested readers can find more information in \cite{Karrer2011}. \\

@@

We will use two real-world networks. The first real-world network is an on-line friendship network among Finnish users of the social music listening and sharing website Last.fm (data from $2006$). Only the largest connected component of this on-line social network is included \cite{ICON}. This data set was originally used in \cite{Toivonen2009}. This on-line social network has size $N = 8003$ nodes; number of edges: $16824$; $Q = 0.79$; $\left<cc\right> = 0.31$; $\left<k\right> = 4.2$. The degree distribution is heterogeneous and is best fitted by a log-normal

@@

The second real-world network is the giant component of the on-line social network of users of the Pretty-Good-Privacy (PGP) algorithm for secure information interchange (data from $2004$) \cite{Boguna2004}\cite{ICON}. This network has $N = 10680$ nodes; number of edges: $24340$; $Q = 0.85$; $\left<cc\right> = 0.27$; $\left<k\right> = 4.56$. The network has a heterogeneous degree distribution, which is best fitted by two power-laws

@@

\begin{figure}[H]
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{}
    \includegraphics[width=8cm, height=7cm]{lastfm_degree_distribution_8x7.png}
    \label{deg_distr_lastfm}
  \end{subfigure}
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{}
    \includegraphics[width=8cm, height=7cm]{PGP_degree_distribution_8x7.png}
    \label{deg_distr_pgp}
  \end{subfigure}
  \captionsetup{format=plain}
  \caption[Degree distribution $P(k)$ of the two real-world networks used in this thesis.]{Degree distribution $P(k)$ of the two real-world networks used in this thesis. \textbf{(a)} Last.fm on-line social network with a log-normal fit ($\mu = 1.008$, $\sigma = 0.95$); \textbf{(b)} PGP on-line trust network with two power-law fits ($a = 0.39$, $\gamma = 1.54$ for $k < 20$ and $a = 5.59$, $\gamma = 2.49$ for $k \geqslant 20$). Curves are fitted to the data with \textsc{scipy.optimize.curve$\_$fit}.}
\label{deg_distr_real_network}
\end{figure}
\noindent
One of the decisive properties of the real-world networks that are used, is their size. The main, and actually only, reason is the computation time. If the networks are too large, it takes weeks, sometimes even months before there would be results. The size of the network models that are implemented, is around a thousand nodes. For the real-world networks, we also use networks with a size of magnitude of thousand to ten-thousand. This benefits the computation time. 
\newpage
\section{Temporal networks}

@@

Binary models are models where each actor can have one of two opinions A or B. A binary model reflects real-life where there are two competing opinions, such as a vote in an election system with two competing candidates or choosing between Windows and Linux \cite{Nguyen2020}.\\
\newline
In this section, the most important binary models will be discussed.

@@

Most of the standard opinion dynamics models use a static network in which actors are always active and exchanging opinions. In these models the opinion dynamics might go as follows: each time step a random actor is selected after which its opinion is updated according to the opinion dynamics model that is implemented. Another possibility is updating all actors at the same time during each time step. \\
\newline
It might be more realistic if actors become active/inactive over time. This would be a better representation of social media platforms, where users are active for some periods of time and then become inactive again. This activation mechanism of the actors gives the network a temporal character. The underlying network structure does not change (each actor has a fixed number of neighbors that does not change over time; no new friends are made or, conversely, no old friends are being unfriended). However, by making actors active and inactive the exchange of information between actors becomes time dependent on which actors are active or not. Hence, this does affect the network structure in some way, because if an actor, for example, does not become active for a long period of time, it does not participate in the exchange of information and the network structure behaves like if this actor is absent. This may affect the dynamics, since it may seem as if the network has a different average degree and some properties like connectivity and clustering may behave differently. \\
\newline
In this thesis, each actor becomes active with a probability $p_i$ every time step. The activation probability is $p = 0.1$, which is constant for all the actors and is chosen for simplicity.
